Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-03 doc/mc-paper.tex
Progress on paper

\newcommand{\pand}{\mathbin{\land^{\sf v}}}
\newcommand{\por}{\mathbin{\lor^{\sf a}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

@@

\newcommand{\assign}{\alpha}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

@@

@@

We have developed a program that can generate POG representations and
equivalence proofs from the dDNNF graphs generated by a leading
knowledge compiler.  It scales to all but the largest formulas from a
recent model counting competition.  We also present two proof
checkers---a fast checker that can handle the large proofs generated,
and a formally verified checker written in the \lean{} proof system.
\end{abstract}

@@

@@

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an {\em assignment} of truth values to the variables, where $0$
  denotes false and $1$ denotes true, i.e., $\assign \colon \varset
  \rightarrow \{0,1\}$.  The set of all possible assignments over
  $\varset$ is denoted $\uassign$.  For each variable $x \in \varset$,
  we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item A variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}

  We highlight the following special classes of Boolean formulas:
  \begin{itemize}
    \item In a {\em normal form} formula, negation is 
      applied only to variables
    \item A normal-form formula is in {\em conjunctive normal form} (CNF) if disjunction is applied only to literals.
    \item A {\em partitioned-operation formula} (or simply a ``partitioned formula'') satisfies the following restrictions on its conjunction and disjunction operations:
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\phi = \bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\phi = \bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned formula.
  \end{itemize}

@@

We propose a general framework for describing a variety of tasks that
summarize some aspect of Boolean formulas.

\begin{definition}[Commutative Ring]
  A {\em commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$, 
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  {\em additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}

\begin{definition}[Ring Evaluation Problem (REP)]
  A {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We define $w(\obar{x})$ to equal $1-w(x)$.

  For Boolean formula $\phi$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem can be cast as an REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow independent
assignments of weights $W$ to the variables and their negations, such
that $W(x) + W(\obar{x}) \not = 1$.  We can convert this into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing algebraic evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, let $\ring$ be a
finite  field with $|\dset| \geq 2 |\varset|$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
least $\left(1-\frac{1}{z}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a partitioned formula:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following:
\begin{description}
\item[Negation] $\rep(\boolnot \phi) = \mulident - \rep(\phi)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i\right) = \prod_{i=1,k} \rep(\phi_i)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i\right) = \sum_{i=1,k} \rep(\phi_i)$
\end{description}
\end{proposition}

@@

As can be seen, a POG is simply a way to represent a partitioned
formula with a sharing of common subformulas, and so we write
$\phi(P)$ as the formula denoted by $P$.

@@

\item It allows arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node in a d-DNNF can have at most two children, and there must be a variable $x$ such that
  the satisfying assignments for one child have $x=1$ while those for the
  other child have $x=0$.  Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{The CRAT Proof System}

The CRAT clausal proof framework provides a means to express a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a POG\@, thus enabling certified
knowledge compilation.  It is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as its core method for
determining that a clause is implication-redundant with respect to other clauses.

Two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em compiler} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
  This compiler might start with the output of another knowledge compiler.
\item The {\em checker} ensures that all of the proof conditions are satisfied.  
\end{itemize}

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
\begin{itemize}
\item 
  Each such declaration implicitly adds an extension variable representing that operation and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for generating extension variables or adding blocked clauses to the proof.
\item The checker must track the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\item Boolean negation is provided implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\end{itemize}
\item
  Assertion clauses can can added when they satisfy the RUP
  property, with respect to a sequence of existing clauses.  This set must be provided explicitly as the {\em hints} for the assertion.
\item
  Deleting clauses requires proving that the resulting formula is not weaker.
  That is, it should not allow any additional satisfying assignment.
  Only the input clauses and those declared by an assertion can be deleted.
      The deletion must be accompanied by a sequence of remaining clauses
      providing a RUP proof of the clause.  This sequence must be provided as the {\em hints} for the deletion.
\end{itemize}

@@

Observe that the defining clauses for the conjunction and disjunction
operations provide a full Tseitin encoding of the operations.  Letting
$\exvarset$ denote the set of added extension variables,  any
assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of
steps that both add and delete clauses.  Each addition must be truth
preserving, that is, any satisfying assignment for the set of clauses
before clause addition should still be a satisfying assignment afterwards.
Each deletion must be falsehood preserving.  That is, there can be no
new satisfying assignments when the clause is deleted.

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\end{itemize}
The sequence of clause addition steps provide a proof that any
assignment $\assign$ to the input variables that satisfies all of the
clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
each proof step that deletes an input clause proves that any
assignment $\alpha$ that falsifies the clause must, when extended,
assign $\assign(r) = 0$.

@@

(E) Assertions\\[1.2ex]

@@

\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & \multicolumn{2}{l}{Hints} & Explanation \\

@@

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
any assignment $\assign$
satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that node $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_10) = 0$.  But this contradicts clause 22,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any assignment that falsifies the clause
must assign $s_{10} = 0$.

@@

Observe that this proof must ``visit'' nodes
$p_5$ and $p_6$ twice, separately considering assignments where $\assign(x_1) = 0$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
general, a naive proof must expand the POG into a tree.
We address this shortcoming by introducing {\em lemmas}, as is described in Section~\ref{sect:lemma}

\section{Generating CRAT from dDNNF}

>>>>

\newcommand{\pand}{\mathbin{\land^{\sf p}}}
\newcommand{\por}{\mathbin{\lor^{\sf p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf v}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf a}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

@@

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

@@

\newcommand{\simplify}[2]{#1|_{#2}}
\newcommand{\expand}{\mathit{Expand}}

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{d4}}
\newcommand{\Dfour}{\progname{d4}}
\newcommand{\generator}{\progname{d2p}}
\newcommand{\Generator}{\progname{d2p}}
\newcommand{\checker}{\progname{crat-check}}
\newcommand{\Checker}{\progname{Crat-check}}

@@

We have developed a program that can generate POG representations and
equivalence proofs from the d-DNNF graphs generated by a leading
knowledge compiler.  It scales to all but the largest formulas from a
recent model counting competition.  We also present two proof
checkers---a fast checker that can handle the large proofs generated,
and a formally verified checker written in the \lean{} proof system.
\end{abstract}

@@

\begin{figure}
\centering{
\begin{tikzpicture}[scale=0.18]
  \draw (8,10) rectangle (20,18);
  \node at (14,15.2) {Knowledge};
  \node at (14,12.8) {Compiler};

  \draw (28,10) rectangle (40,18);
  \node at (34,15.2) {Proof};
  \node at (34,12.8) {Generator};

  \draw (48,10) rectangle (60,18);
  \node at (54,15.2) {Proof};
  \node at (54,12.8) {Checker};

  \draw (48,0) rectangle (60,8);
  \node at (54,5.2) {Ring};
  \node at (54,2.8) {Evaluator};

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (48,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\phi$}};
  \node [above] at (2,20) {\texttt{.cnf}};

  %% NNF
  \draw[thick] (20,12) [-latex] -- (28,12);
  \node [above] at (24,12) {\texttt{.nnf}};

  %% CRAT
  \draw[thick] (40,12) [-latex] -- (48,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (48,4);
  \node [above] at (44,12) {\texttt{.crat}};

  %% OK/Not
  \draw[thick] (60,14) [-latex] -- (64,14) ;
  \node [right] at (64,15.2) { OK /} ;
  \node [right] at (64,12.8) { Not OK} ;
  \draw[thick] (60,4) [-latex] -- (64,4) ;
  \node [right] at (64,4) {\large {$\textbf{R}(\phi)$}};

\end{tikzpicture}
}
\caption{Tool Chain. The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

@@

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an {\em assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
  \subseteq \varset$.  We say the assignment is {\em total} when it
  assigns a value to every variable ($\varset' = \varset$), and that
  it is {\em partial} otherwise.
  The set of all possible total assignments over
  $\varset$ is denoted $\uassign$.

For each variable $x \in \varset$,
  we define {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em conjunction} $\phi = \bigwedge_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{i=1,k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em disjunction} $\phi = \bigvee_{i=1,k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{i=1,k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{i=1,k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight the following special classes of Boolean formulas:
  \begin{description}
    \item[Normal form:] A formula where negation is applied only to variables.
    \item[Conjunctive Normal Form (CNF):] A normal form formula where  disjunction is applied only to literals.
    \item[Partitioned-Operation:] A formula satisfying the following for all conjunction and disjunction operations: 
      \begin{itemize}
      \item The arguments to a conjunction must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1,k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to a disjunction must have disjoint models.  That is, operation
        $\bigvee_{i=1,k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{itemize}
  \end{description}
     We let $\pand$ and $\por$ denote the conjunction and disjunction operations in a partitioned-operation formula.
     A formula in conjunctive normal form can be represented by a set
     of {\em clauses}, each of which is a set of literals.  Each
     clause then represents the disjunction of the literals, and the formula is then the conjunction of its clauses.

@@

We propose a general framework for 
summarizing properties of Boolean formulas.

\begin{definition}[Commutative Ring]
  A {\em commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$, 
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  {\em additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}
We write $a - b$ as a shorthand for $a + -b$.

\begin{definition}[Ring Evaluation Problem (REP)]
  For ring $\ring$, a {\em ring labeling} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x})$ to equal $1-w(x)$.

  For Boolean formula $\phi$, the {\em ring evaluation problem} (REP) is to compute:
  \begin{eqnarray}
    \rep(\phi) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
\label{def:labeling}
\end{definition}

A wide variety of important tasks for Boolean functions can be
expressed as ring evaluation problems.  The 
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$
It can be cast as an REP by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.  
Ring evaluation of formula $\phi$ will then compute the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$ will then
yield the number of models.

The {\em weighted model counting} (WMC) problem is also defined over
rational numbers.  Some formulations of WMC allow 
assigning weights $W$ to both the variables and their negations, such
that there can be variables $x$ with $W(x) + W(\obar{x}) \not = 1$.  We can convert this more general form into a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with label $w(x) = W(x)/r(x)$ for each
variable $x$, and then computing the final result as $\rep(\phi)\;
\rmul\; \prod_{x \in \varset} r(x)$.  Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a randomized test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1) = \rep(\phi_2)$
if they are logically equivalent.  If they are not equivalent, then
the probability that $\rep(\phi_1) \not = \rep(\phi_2)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Therefore, ring evaluation can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

Possible generalizations of this framework are discussed in Section~\ref{sect:future}.

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following:
\begin{description}
\item[Negation] $\rep(\boolnot \phi) = \mulident - \rep(\phi)$
\item[Partitioned Conjunction] $\rep\left(\Pand_{i=1,k} \phi_i\right) = \prod_{i=1,k} \rep(\phi_i)$
\item[Partitioned Disjunction] $\rep\left(\Por_{i=1,k} \phi_i\right) = \sum_{i=1,k} \rep(\phi_i)$
\end{description}
\end{proposition}

@@

As can be seen, a POG is simply a way to represent a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node the graph can be viewed as a partitioned-operation formula, and so we write
$\phi(u)$ as the formula denoted by node $u$ and write $\phi(P) = \phi(r)$ for root $r$ of $P$.

@@

\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  disjunction node $v$ in a d-DNNF can have at most two children $u_1$ and $u_0$ for which there must be a {\em splitting variable} $x$ such that
  any total assignment $\assign$ satisfying $\phi(u_1)$ has $\assign(x)=1$, while any total assignment satisfying $\phi(u_0)$ has
  $\assign(x)=0$.  Our generalization is provided to allow an
  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} into
  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  Both of these generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{The CRAT POG Representation and Proof System}

A CRAT file provides both a declaration of a POG, as well as a checkable, clausal
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.  The proof system is based on extended
resolution~\cite{Tseitin:1983} with reverse unit propagation
(RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
proving that a set of clauses logically implies another clause.

As is shown in Figure~\ref{fig:chain}
two programs are involved in producing a certified compilation of a formula:
\begin{itemize}
\item The {\em proof generator} generates a CRAT file that both describes a POG and provides a proof that this POG is equivalent to the input formula.
  This compiler might start with the output of another knowledge compiler.
\item The {\em proof checker} ensures that all of the proof conditions are satisfied.  
\end{itemize}

The CRAT format draws its inspiration from the LRAT format for Boolean
formulas~\cite{lrat} and the QRAT format for quantified Boolean
formulas (QBF)~\cite{heule:JAR2014}.  The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Each such declaration implicitly adds an extension variable representing the POG node and a set of {\em defining} clauses
  encoding the conjunction or disjunction operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed to avoid requiring
  $k\,(k-1)/2$ proofs of disjointness for larger values of $k$.
\end{itemize}

@@

Observe that the defining clauses for the conjunction and disjunction
operations provide a full Tseitin encoding of the operations.  Letting
$\exvarset$ denote the set of added extension variables,  any
total assignment $\assign\colon \varset \rightarrow \{0,1\}$ has a unique extension,
which we also call $\assign$, with $\assign\colon (\varset \cup \exvarset) \rightarrow \{0,1\}$.

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of
steps that both add and delete clauses.  Each addition must be truth
preserving, that is, any satisfying total assignment for the set of clauses
before clause addition should still be a satisfying assignment afterwards.
Each deletion must be falsehood preserving.  That is, there can be no
new satisfying assignments when the clause is deleted.

At the completion of the proof, the following conditions must hold:
\begin{itemize}
\item All of the input clauses have been deleted.
\item There should be exactly one clause left that was added via RUP
  addition, and this should be a unit clause consisting of the root literal $r$.
\end{itemize}
The sequence of clause addition steps provide a proof that any
total assignment $\assign$ to the input variables that satisfies all of the
clauses must, when extended, assign $\assign(r) = 1$.  Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must, when extended,
assign $\assign(r) = 0$.

@@

(E) CRAT Assertions\\[1.2ex]

@@

\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hints} & Explanation \\

@@

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $s_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must extend to one with $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that node $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hints for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause, but assigns $\assign(s_{10}) = 1$.
proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hints then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 forces
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 22,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $s_{10} = 0$.

@@

Observe that this proof must ``visit'' nodes
$p_5$ and $p_6$ twice, separately considering total assignments where $\assign(x_1) = 0$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  In
general, a naive proof must expand the POG into a tree.
We address this shortcoming by introducing {\em lemmas}, as is described in Section~\ref{sect:lemma}

\section{Generating CRAT from d-DNNF}

Figure~\ref{fig:chain} illustrates one method of generating a CRAT
file from an input formula in conjunctive normal form.  Our tool chain
is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
but it would also work for other tools that use a top-down search
strategy to generate a d-DDNNF representation.  Converting a d-DNNF
graph into a POG is straightforward, each node in a d-DNNF graph can
be directly represented as a POG node.  We perform simplifications on
the d-DNNF representation to eliminate any Boolean constants.  We can
therefore assume that the POG contains only conjunction, disjunction,
and literal nodes.

\section{CNF Implies POG}

The most challenging part of the proof is to show that any total
assignment $\alpha$ that is a model of $\phi$ will, when extended,
have $\assign(r) = 1$ for root literal
$r$.  This part of the proof consists of a series of assertion steps
leading to one adding $r$ as a unit clause.

In the following, we describe proof generation as a process that
recursively traverses the graph from the root downward and
generates proof steps from the lower nodes upward.  This process
effectively expands the POG into a tree and therefore can be very
inefficient.  We eliminate this inefficiency with the addition of
lemmas, as is described later.
